The impact of the COVID-19 pandemic on academic productivity

### 2022-06-22 article/ms.tex
returned to rsos

%%%% Abstract text to be placed here %%%%%%%%%%%%
\begin{abstract}
`Publish or perish’ is an expression describing the pressure on academics to consistently publish research to ensure a successful career in academia. 
With a global pandemic that has changed the world, how has it changed academic productivity? 
Here we show that academics are posting more papers on the arXiv pre-print server than would be expected had the pandemic not occurred: 341,474 pre-prints were posted between 2020-2021, up 14\% from 2019 and above the predictions of our Gaussian process model.
%168,630 were posted in 2020 , a +12.6\% change from 2019 and $+1.4\sigma$ deviation above the 162,577 predicted by our Gaussian process model. % todo: reword
We find that that most individual fields were not measurably affected by the pandemic.
However, we do observe a significant ongoing decline in \add{general} physics pre-prints, as well as temporary declines in areas of high-energy physics, where the cancellation of a key conference played a key role.
The most significant increase occurred in quantitative biology where a 50\%~increase ($+8\sigma$) in pre-prints posted in 2020 was driven by research related to the COVID-19 pandemic. 
Some of these `COVID papers' are by biologists using arXiv for the first time, but many are written by researchers from other fields (e.g., physicists, mathematicians). 
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The exception is \add{general physics (\arxiv\ category \texttt{physics}; not all physics-related domains}), which shows the largest decline in pre-prints by number since the pandemic began. In 2020 the number posted to general physics actually exceeded our model predictions (15,242 compared to 14,746) by $1\sigma$, driven by a short peak of pre-prints posted to Physics and Society (\texttt{physics.soc-ph}) in early 2020. When we segment the number of pre-prints posted to physics by their sub-category we find that Computational Physics (\texttt{physics.comp-ph}) and Applied Physics (\texttt{physics.app-ph}) had significant drops in pre-print counts since the pandemic began (Figure~\ref{fig:physics}). All other physics sub-categories remained largely stable with time. The category of mathetmatical physics (\texttt{math-ph}) also showed a decline in pre-prints since the pandemic began, but this category has about an order of magnitude fewer pre-prints posted than physics. The mathematical physics category does not have defined sub-categories, making it more difficult to evaluate whether this decline occurred in particular areas of study.

@@

The COVID-19 pandemic appears to have had minimal impact on collective research productivity to date.
However, some significant changes were noted within specific areas of research. The largest persistent decline in pre-prints has occurred in \add{general physics (arXiv primary category \texttt{physics}), largely due to a decline in  pre-prints posted to computational and applied physics areas within the \texttt{physics} primary category. As of mid-2022, this steady decline continues.} Amongst individual fields the largest \emph{relative} decline in a single field is explainable by a high energy lattice physics conference that was cancelled in 2020. In this study we have only focussed on the quantity of publications as a measure for productivity, and not their quality. While citations are the most used measure of impact of academic publications, that metric becomes a more biased statistic when many related pre-prints are all being posted nearly at the same time\cite{Fassin:2021}. The relatively long timescales of academic research would suggest that the full impact on academic productivity due to the COVID-19 pandemic is yet to be seen. The methodology and results presented here provide a quantitative framework for future investigations of the longer term impacts of the pandemic.

@@

\aucontribute{AC and IM conceived the idea to examine changes in academic productivity due to the pandemic. \add{PR} and IM helped guide the investigation and provided ideas on new avenues to explore and tests to perform. AC conducted the analysis. AC led the writing of the manuscript, with all authors contributing significantly. All authors read and iterated upon the text of the final manuscript.}

@@

\bibitem{Viglione:2020}
{Viglione, G.}, {Are women publishing less during the pandemic? Here\add{'}s what the data say}. \emph{Nature}, \textbf{581}, 356-366, doi:10.1038/d41586-020-01294-9 \newblock (2020).

@@

\add{\bibitem{ZenodoRecord} Casey, Andrew R. (2022). Impact of the COVID-19 pandemic on academic productivity [Data set]. Zenodo. https://doi.org/10.5281/zenodo.6615684}

>>>>

%%%% Abstract text to be placed here %%%%%%%%%%%%
\begin{abstract}
`Publish or perish’ is an expression describing the pressure on academics to consistently publish research to ensure a successful career in academia. 
With a global pandemic that has changed the world, how has it changed academic productivity? 
Here we show that academics are posting more papers on the arXiv pre-print server than would be expected had the pandemic not occurred: 341,474 pre-prints were posted between 2020-2021, up 14\% from 2019 and above the predictions of our Gaussian process model.
%168,630 were posted in 2020 , a +12.6\% change from 2019 and $+1.4\sigma$ deviation above the 162,577 predicted by our Gaussian process model. % todo: reword
We find that that most individual fields were not measurably affected by the pandemic.
However, we do observe a significant ongoing decline in general physics pre-prints, as well as temporary declines in areas of high-energy physics, where the cancellation of a key conference played a key role.
The most significant increase occurred in quantitative biology where a 50\%~increase ($+8\sigma$) in pre-prints posted in 2020 was driven by research related to the COVID-19 pandemic. 
Some of these `COVID papers' are by biologists using arXiv for the first time, but many are written by researchers from other fields (e.g., physicists, mathematicians). 
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The exception is general physics (\arxiv\ category \texttt{physics}; not all physics-related domains), which shows the largest decline in pre-prints by number since the pandemic began. In 2020 the number posted to general physics actually exceeded our model predictions (15,242 compared to 14,746) by $1\sigma$, driven by a short peak of pre-prints posted to Physics and Society (\texttt{physics.soc-ph}) in early 2020. When we segment the number of pre-prints posted to physics by their sub-category we find that Computational Physics (\texttt{physics.comp-ph}) and Applied Physics (\texttt{physics.app-ph}) had significant drops in pre-print counts since the pandemic began (Figure~\ref{fig:physics}). All other physics sub-categories remained largely stable with time. The category of mathetmatical physics (\texttt{math-ph}) also showed a decline in pre-prints since the pandemic began, but this category has about an order of magnitude fewer pre-prints posted than physics. The mathematical physics category does not have defined sub-categories, making it more difficult to evaluate whether this decline occurred in particular areas of study.

@@

The COVID-19 pandemic appears to have had minimal impact on collective research productivity to date.
However, some significant changes were noted within specific areas of research. The largest persistent decline in pre-prints has occurred in {general physics (arXiv primary category \texttt{physics}), largely due to a decline in  pre-prints posted to computational and applied physics areas within the \texttt{physics} primary category. As of mid-2022, this steady decline continues.} Amongst individual fields the largest \emph{relative} decline in a single field is explainable by a high energy lattice physics conference that was cancelled in 2020. In this study we have only focussed on the quantity of publications as a measure for productivity, and not their quality. While citations are the most used measure of impact of academic publications, that metric becomes a more biased statistic when many related pre-prints are all being posted nearly at the same time\cite{Fassin:2021}. The relatively long timescales of academic research would suggest that the full impact on academic productivity due to the COVID-19 pandemic is yet to be seen. The methodology and results presented here provide a quantitative framework for future investigations of the longer term impacts of the pandemic.

@@

\aucontribute{AC and IM conceived the idea to examine changes in academic productivity due to the pandemic. {PR} and IM helped guide the investigation and provided ideas on new avenues to explore and tests to perform. AC conducted the analysis. AC led the writing of the manuscript, with all authors contributing significantly. All authors read and iterated upon the text of the final manuscript.}

@@

\bibitem{Viglione:2020}
{Viglione, G.}, {Are women publishing less during the pandemic? Here's what the data say}. \emph{Nature}, \textbf{581}, 356-366, doi:10.1038/d41586-020-01294-9 \newblock (2020).

@@

\bibitem{ZenodoRecord} Casey, Andrew R. (2022). Impact of the COVID-19 pandemic on academic productivity [Data set]. Zenodo. https://doi.org/10.5281/zenodo.6615684
